---
title: "Task 1: Data Wrangling & Visualization of Sierra Amphibians"
author: "Samantha Kuglen"
date: "1/22/2022"
output: html_document
---
### A. Overview
This report provides an exploratory review of amphibian abundance data in the Sierra Lakes recorded from 1995 - 2002. The purpose fo this analysis is to wrangle and visualize data for mountain yellow-legged frogs (*Rana muscosa*). This data is important because it has supported research and management of Sierra Nevada aquatic ecosystems and their terrestrial interfaces. 

### B. Data and Analysis

This amphibian data contains observations physical characteristics of and surveyed aquatic communities for >8,000 lentic water bodies in the southern Sierra Nevada, including lakes, ponds, marshes, and meadows. The data was collected and made available by the [Sierra Lakes Inventory Project (SLIP)](https://portal.edirepository.org/nis/mapbrowse?packageid=edi.577.2). All analyses are in R version 4.0.2 using RStudio version 1.3.1056.

```{r setup, include=TRUE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

# Attach packages
library(tidyverse)
library(here)
library(lubridate)
library(patchwork)
library(paletteer)
```

```{r}
# Read in the data
amphibian_data <- read_csv(here("data", "sierra_amphibians.csv"))
```

#### I. Data Wrangling and Visualization I  
Find total number of adult, subadult, and tadpole mountain yellow-legged frogs observed in the entire study by life stage and year, excluding Eggmass level, then store results for visualization. 
```{r}
# Filter for *Rana muscosa* species only, exclude EggMass and use `lubridate` package to retrieve year for grouping 
total_yellow_frogs <- amphibian_data %>% 
  filter(amphibian_species == "RAMU", amphibian_life_stage != "EggMass") %>% 
  mutate(survey_date = mdy(survey_date), year = year(survey_date)) 

# Visualize total counts across all water bodies, grouping by amphibian_life_stage 
plot1 <- ggplot(data = total_yellow_frogs, aes(x = year, group = amphibian_life_stage, fill = amphibian_life_stage)) +
  geom_bar(position = "dodge") + # separates counts
  scale_fill_manual(values = c("#b6fc03", "#03a9fc", "#fc8803")) + 
  labs(title = "Total Mountain Yellow-Legged Frogs Per Year",
       subtitle = "Sierra Lakes (1995 - 2002)",
       x = "Year",
       y = "Count",
       fill = "Amphibian Life Stage") +
  theme_linedraw() + 
  theme(plot.title = element_text(hjust = 0.5, face = "bold"), # center title/subtitle
        plot.subtitle = element_text(hjust = 0.5),
        plot.background = element_rect(fill = "#D3D3D3"),
        legend.key = element_rect(fill = "black"),
        legend.background = element_rect(fill = "#f5ebeb")) +
  scale_x_continuous(breaks = c(1995:2002)) + # add breaks to show all years 
  scale_fill_paletteer_d(palette = "colorBlindness::Blue2Gray8Steps") # colorblind-friendly palette

plot1 + scale_alpha(guide = "none") # suppress duplicate legend
```

#### II. Data Wrangling and Visualization II
Find the 5 lakes with greatest total counts of yellow frogs, with Lake ID as categorical label on the x-axis and total yellow-legged frog counts on the y-axis, then store results for visualization.

```{r}
top_5_lakes <- total_yellow_frogs %>% 
  filter(amphibian_life_stage != "Tadpole") %>% 
  group_by(lake_id) %>% 
  summarize(total = sum(amphibian_number)) %>% 
  slice_max(order_by = total, n = 5) 

#as.character change to read as a string because the ID is a number

top_5_lakes$lake_id <- sapply(top_5_lakes$lake_id, function(x) paste("Lake", x))
top_5_lakes$lake_id <- fct_reorder(as.character(top_5_lakes$lake_id), top_5_lakes$total)

plot2 <- ggplot(data = top_5_lakes, aes(y = lake_id, x = total), aes(fill = total)) + 
  geom_col(aes(fill = lake_id)) +
  labs(title = "Lakes With Greatest Adult and SubAdult Yellow-Legged Frogs",
       subtitle = "Sierra Lakes (1995 - 2002)",
       x = "Total Count", 
       y = "Lake ID") +
  theme_linedraw() + 
  theme(plot.title = element_text(hjust = 0.5, face = "bold"), # center title/subtitle
        plot.subtitle = element_text(hjust = 0.5),
        plot.background = element_rect(fill = "#D3D3D3"),
        axis.title.y = element_text(margin = margin(r = 20)), # add space btwn y-axis label and graph
        legend.position = 'none')  + # hide legend (unnecessary)
    scale_fill_paletteer_d(palette = "colorBlindness::Green2Magenta16Steps") # colorblind-friendly palette

```

### C. Final Data Visualization
Combine plots into a single compound figure
```{r}
# Use `patchwork` package syntax to place graphs on top of each other
plot1 / plot2
```

**Figure 1.** 
