---
title: "Task 2: Model Selection of CalCOFI Seawater Sample Data "
author: "Samantha Kuglen"
date: "1/23/2022"
output: html_document
---
### A. Overview
This report provides an exploratory review on the relationship between O2 saturation of seawater off California’s coast and several physical and chemical variables. Specifically, we examine and compare (1) Oxygen saturation as a function of water temperature, salinity, and phosphate concentration and (2) Oxygen saturation as a function of water temp, salinity, phosphate concentration, and depth.The purpose of this analysis is to determine which of the two models is more parsimonious, or in other words, provides predictive performance with simplicity and accuracy. This data is important because it provides important nutrient, salinity, and temperature information about the California Current System. 

### B. Data and Analysis
The data contain observations from [CalCOFI seawater sample data](https://calcofi.org/ccdata.html). A 70+ year hydrographic time-series includes temperature, salinity, oxygen and phosphate observations, along with nutrient analysis of silicate, nitrate and nitrite. An AIC comparison between the two models is used, to inform which model is a better fit, with preference for the model with lower AIC. Following AIC comparison of two linear regression models, a ten-fold cross validation is performed. The cross validation reserves a subset of data (test data) and trains our model using the rest (training data) to estimate the model parameters. All analyses are in R version 4.0.2 using RStudio version 1.3.1056.

```{r setup, include=TRUE, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

# Attach packages
library(tidyverse)
library(here)
library(AICcmodavg)
library(kableExtra)
library(equatiomatic)

```

```{r}
# Read in the data
calcofi_data <- read_csv(here("data", "calcofi_seawater_samples.csv"))
```

### I. AIC Comparison of Two Linear Regression Models 
**Model 1 (`mdl1`):** Oxygen saturation as a function of water temperature, salinity, and phosphate concentration. **Model 2 (`mdl2`):** Oxygen saturation as a function of water temp, salinity, phosphate concentration, and depth.

```{r}
f1 <- o2sat ~ t_deg_c + salinity + po4u_m
mdl1 <- lm(f1, data = calcofi_data)
  
f2 <- o2sat ~ t_deg_c + salinity + po4u_m + depth_m
mdl2 <- lm(f2, data = calcofi_data)
```

AICc is used to select the better model, with a consideration the difference of AIC between the two models. 

```{r}
aic_table <- AICcmodavg::aictab(list(mdl1, mdl2)) %>% 
  rename("Model" = "Modnames") 


kable(aic_table,
      caption = "Table 1: Models 1 and 2 AIC corrected (AICc) Values") %>% 
  kable_styling(bootstrap_options = "striped", 
                full_width = FALSE) 

```

Model 2's AICc is `r round(aic_table[1, 3], 3)` while Model 1's AICc `r round(aic_table[2, 3], 3)`, with a difference in AIC's of `r round(aic_table[2, 3], 3) - round(aic_table[1, 3], 3)`. Model 2's AICc is lower than model 1 and means that it is the preferred model and a better fit, based on AICc. 

### II. Ten-fold Cross Validation of the Two Models
Ten-fold cross validation is used with root-mean-square error (RMSE) as the scoring method and the final model is trained on the full dataset. 
```{r}
folds <- 10
fold_vec <- rep(1:folds, length.out = nrow(calcofi_data))

set.seed(50)

seawater_fold <- calcofi_data %>% 
  mutate(group = sample(fold_vec, size = n(), replace = FALSE))

# First fold
test_df <- seawater_fold %>% 
  filter(group == 1) 
train_df <- seawater_fold %>% 
  filter(group != 1)
```

RMSE goes in reverse order of these operations - find error (predicted actual), square it, find the average, then take the square root. 

```{r}
calc_rmse <- function(x, y) {
  rmse_result <- (x - y)^2 %>%  mean() %>%  sqrt()
  return(rmse_result)
}
```

The training dataset is used to create two linear regression models, based on the formulas above.
```{r}
training_mdl1 <- lm(f1, data = train_df)
training_mdl2 <- lm(f2, data = train_df)
```

The trained models are used to predict on test data. 
```{r}
predict_test <- test_df %>% 
  mutate(model1 = predict(training_mdl1, test_df),
         model2 = predict(training_mdl2, test_df))

rmse_predict_test <- predict_test %>% 
  summarize(rmse_mdl1 = round(calc_rmse(model1, o2sat), digits = 3),
            rmse_mdl2 = round(calc_rmse(model2, o2sat), digits = 3)) 

kable(rmse_predict_test,
      col = c("RMSE Model 1",
              "RMSE Model 2"),
      caption = "Table 2: RMSE Values for Models 1 and 2") %>% 
  kable_styling(bootstrap_options = "striped", 
                full_width = FALSE)

```

Calculation over all folds and taking the average of RMSE values. 
```{r}
rmse_df <- data.frame()

for(i in 1:folds) {
  kfold_test_df <- seawater_fold %>% 
    filter(group == i)
  kfold_train_df <- seawater_fold %>% 
    filter(group != i)
  
  kfold_mdl1 <- lm(f1, data = kfold_train_df)
  kfold_mdl2 <- lm(f2, data = kfold_train_df)
  
  kfold_pred_df <- kfold_test_df %>% 
    mutate(mdl1 = predict(kfold_mdl1, kfold_test_df),
           mdl2 = predict(kfold_mdl2, .))
  
  kfold_rmse <- kfold_pred_df %>% 
    summarize(rmse_mdl1 = calc_rmse(mdl1, o2sat),
              rmse_mdl2 = calc_rmse(mdl2, o2sat))
  
  rmse_df <- bind_rows(rmse_df, kfold_rmse)
}

rmse_table <- rmse_df %>% 
  summarize(mean_rmse_mdl1 = mean(rmse_mdl1),
            mean_rmse_mdl2 = mean(rmse_mdl2)) 

kable(rmse_table,
      col = c("Mean RMSE Model 1",
                      "Mean RMSE Model 2"),
      caption = "Table 3: Average RMSE Values for Models 1 and 2") %>% 
  kable_styling(bootstrap_options = "striped", 
                full_width = FALSE)
```
We find that Model 2 has a lower average RMSE than Model 1, at values of `r round(rmse_table[1, 2], 3)` and `r round(rmse_table[1, 1], 3)`, respectively. The 10 cross validation informs us that Model 2 has the best fit and predicts the observed values for oxygen saturation `o2sat` with the most accuracy. 

Once we have chosen the model via cross validation (Model 2), we train our final model on our entire dataset. 
```{r}
final_mdl <- lm(f2, data = calcofi_data)
```
Our final parameterized model: 
`r equatiomatic::extract_eq(final_mdl, wrap = TRUE)`

Our final parameterized model with coefficients: 
`r equatiomatic::extract_eq(final_mdl, wrap = TRUE, use_coefs = TRUE)`

### Summary
Data analysis reveals the following findings:

- Model 2's AICc is lower than Model 1 and means that it is the preferred model and a better fit, based on AICc.
- Model 2 has a lower average RMSE than Model 1. The 10 cross validation informs us that Model 2 has the best fit and predicts the observed values for oxygen saturation `o2sat` with the most accuracy. 
- Model 2: *Oxygen saturation as a function of water temp, salinity, phosphate concentration, and depth* is the most accurate model.


### Data Citation
CalCOFI data are available for use without restriction. Data downloaded from https://calcofi.org/ccdata.html.  Accessed 1/10/2022.


